---
title: "Group Project"
author: "Jane Zhou and Lena Ivens"
date: "11/23/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Import and Consolidate Data
```{r, cache=TRUE}
# read CSV file containing 5% sample
sample19 <- read.csv("sample19.csv")

# convert fields to factors as needed
sample19$starttime <- as.factor(sample19$starttime)
sample19$start.station.name <- as.factor(sample19$start.station.name)
sample19$end.station.name <- as.factor(sample19$end.station.name)
sample19$usertype <- as.factor(sample19$usertype)
sample19$birth.year <- as.factor(sample19$birth.year)
sample19$gender <- as.factor(sample19$gender)
```

## Visualize Data
```{r}
# load library dplyr (for data manipulation)
library(dplyr)
# load library ggplot2 (for data visualization)
library(ggplot2)
# load library "hans" (for calculating distance)
library(hans)
```

### Bikes
```{r}
# make column that specifies distance per trip in miles
sample19$distance <- haversine(sample19$start.station.latitude, sample19$start.station.longitude, sample19$end.station.latitude, sample19$end.station.longitude)*0.621371

# group sample by bikeid and distance
bikes <- sample19 %>% 
  group_by(bikeid) %>%
  mutate(distance_per_bike = sum(distance)) %>%
  mutate(trips_per_bike = n())

# remove duplicate bikeids and create new data frame
bikes1 <- select(bikes, bikeid, distance_per_bike, trips_per_bike)
bikes1 <- bikes1[!duplicated(bikes1$bikeid), ]

# create a scatter plot with distance and number of trips
plot_a <- ggplot(data = bikes1, aes(x = trips_per_bike, y = distance_per_bike, colour = distance_per_bike)) + geom_point(size=1, alpha=0.5) + scale_colour_gradient(low="greenyellow", high="darkgreen") + theme(legend.position = "none")
plot_a

# identify the most and least used bikes
mostusedbikes <- head(bikes1[order(bikes1$distance_per_bike, decreasing = TRUE),], n = 100)
mostusedbikes
leastusedbikes <- tail(bikes1[order(bikes1$distance_per_bike, decreasing = TRUE),], n = 100)
leastusedbikes
```

- Some bikes have been used more than others, which may be an indicator for Citi Bike to check them for maintenance as the tires could be wearing out. As we can see, the distance traveled ranges from 0 miles to `r round(max(bikes1$distance_per_bike),2)` miles. The scatterplot reveals that there is no unusual pattern since the more a bike is booked the greater the distance traveled. 

```{r}
# create bar plot with num_trips on x_axis
plot_b <- ggplot(data=bikes1, aes(x=trips_per_bike, fill=..count..)) + geom_bar() + scale_fill_gradient(low="greenyellow", high="darkgreen") + theme(legend.position = "none")
plot_b

# find quartiles along with extreme percentiles of num_trips
quantile_num_trips <- quantile(bikes1$trips_per_bike, probs=c(0.01,0.05,0.1,0.25,0.5,0.75,0.9,0.95,0.99,1))
quantile_num_trips
```
- The distribution of the number of trips each bike made in 2019 is skewed right with the mean `r round(mean(bikes1$trips_per_bike),0)` being greater than the median `r quantile_num_trips[5]`.
- Bike `r bikes1$bikeid[bikes1$trips_per_bike == max(bikes1$trips_per_bike)]` has been used the most times with `r max(bikes1$trips_per_bike)` trips made. Although it may not be the bike that traveled the most distance, it should still have traveled a fair amount as there is a strong positive correlation between number of trips and distance traveled (seen in the scatterplot above). One possible reason could be because Bike `r bikes1$bikeid[bikes1$trips_per_bike == max(bikes1$trips_per_bike)]` has always been circulating in the busier parts of NYC where people tend to bike shorter distances but more frequently.
- 1% of the bikes have only been used `r quantile_num_trips[1]` times or less throughout the whole year. This could be because:
  + They have been taken away for maintenance.
  + There is something wrong with them and should be checked for maintenance.
  + They were taken to unpopular stations that tend to have a lot of bike surplus and by chance were rarely selected by bikers to ride. 

### Gender 
```{r}
# change gender category labels (0=unknown; 1=male; 2=female)
sample19$gender_new <- ifelse(sample19$gender == 1, "male", ifelse(sample19$gender == 2, "female", "unknown"))

# create bar plot with gender on x-axis 
plot1 <- ggplot(data=sample19, aes(x=gender_new, fill=..count..)) + geom_bar() + scale_fill_gradient(low="greenyellow", high="darkgreen") + labs(x="gender") + theme(legend.position = "none")
plot1
```

- Almost three times as many males ride Citi Bikes as females.

### Age
```{r}
# calculate bikers' ages and save to new column
sample19$age <- 2019 - as.numeric(as.character(sample19$birth.year))

# create histogram plotting age
plot2 <- ggplot(data=sample19, aes(x=age, fill=..count..)) + geom_histogram(binwidth=1) + scale_fill_gradient(low="greenyellow", high="darkgreen")
plot2
```

- Most Citi Bikers are in their mid-20s to late-40s (the average age is `r round(mean(sample19$age), 0)`).
- Interesting peak at 50. Is exercising more and being healthy a common new year's resolution for 50 year olds?
- The youngest Citi Biker is `r min(sample19$age, na.rm = TRUE)`.
- The oldest Citi Biker is `r max(sample19$age, na.rm = TRUE)`.

### User Type
**FREQUENCY**
```{r}
# create bar plot with user type on x-axis 
plot3 <- ggplot(data=sample19, aes(x=usertype, fill=usertype)) + geom_bar() + scale_fill_manual(values=c("chartreuse3", "darkgreen")) + theme(legend.position = "none")
plot3
```

- There are about six times as many subscribers as regular customers.

**TRIP DURATION**
```{r}
# create violin plot with user type on x-axis and trip duration on y-axis
plot4 <- ggplot(data=sample19, aes(x=usertype, y=tripduration, fill=usertype)) + geom_violin() + scale_y_log10() + scale_fill_manual(values=c("darkgreen", "chartreuse3")) + theme(legend.position = "none")
plot4
```

- Subscribers tend to bike for shorter durations than regular customers, which makes sense because subscribers are probably NYC locals. They can ride Citi Bikes whenever they want, and they ride more for the purpose of getting from one place to another efficiently (unlike tourists who might spend more time biking around leisurely).

### Number of Trips
**OVER A DAY**
```{r}
# extract hour from start time and save to new column
sample19$hour <- substring(sample19$starttime, 12, 13)

# create bar plot with hour on x-axis
plot5 <- ggplot(data=sample19, aes(x=hour, fill=..count..)) + geom_bar() + scale_fill_gradient(low="greenyellow", high="darkgreen")
plot5
```

- The number of trips peak around 8-9am and 5-6pm (rush hours), which suggests that many people use Citi Bike to get to and from school and work.
- The number of trips rises throughout the afternoon perhaps as a result of students getting off from school.

**OVER A WEEK**
```{r}
# extract date from start time, convert to day of the week, and save to new column
sample19$day <- weekdays(as.Date(substring(sample19$starttime, 1, 10))) 

# order the bars
sample19$day <- factor(sample19$day,levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))

# create bar plot with day of the week on x-axis
plot6 <- ggplot(data=sample19, aes(x=day, fill=day)) + geom_bar() + scale_fill_manual(values=c("darkgreen", "darkgreen", "darkgreen", "darkgreen", "darkgreen", "chartreuse3", "chartreuse3")) + theme(legend.position = "none")
plot6
```

- Less people use Citi Bike on weekends, especially on Sundays. This graph supports the hypothesis that the customer base of Citi Bike is primarily composed of workers and students (some jobs and classes are Monday-Saturday). Also, Sunday is usually the day to rest - no work and less errands to run. 

**OVER A YEAR**
```{r}
# extract month from start time and save to new column
sample19$month <- substring(sample19$starttime, 6, 7)

# create bar plot with month on x-axis
plot7 <- ggplot(data=sample19, aes(x=month, fill=..count..)) + geom_bar() + scale_fill_gradient(low="greenyellow", high="darkgreen")
plot7
```

- People use Citi Bike the most during September, perhaps because  of the nice weather and the fact that students have started school again.
- Less people use Citi Bike during the winter, most likely because snow and ice make the sidewalks and roads slippery (safety hazard). There is a sharp decline in trips during November and December, which is generally when it begins snowing in NYC.

## The Average Citi Biker
- male
- young or middle-aged adult
- NYC area resident
- student or worker