---
title: "Group Project"
author: "Wesley Chiu"
date: "11/23/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Set up libraries here
```{r}
library(dplyr)
library(stringr)
library(ggplot2)
library(geosphere)
```

```{r, cache=TRUE}
# 
# This section is for making the files, our group decided to use one common sample file instead
# # First, I downloaded all the data files and extracted them to a folder within my project folder. 
# 
# # Next, I used this code to merge all the files. 
# filenames <- list.files(path="C:/Users/wesle/Desktop/TO404/Final Group Project/MergeFolder", pattern="*.csv")
# fullpath <- file.path("C:/Users/wesle/Desktop/TO404/Final Group Project/MergeFolder",filenames)
# citibike <- do.call("rbind",lapply(fullpath,FUN=function(files){ read.csv(files)}))
# 
# # I then want to take a sample of this data to work with 
# # Set a seed for the random sample so that the data is consistent
# set.seed(42)
# citisample <- sample_frac(citibike, size = .05)
# # Free up memory by getting rid of initial merged file 
# citibike <- NULL

# Read in sample file
citisample <- read.csv("sample19.csv")

# then add weather data to the data frame
weather <- read.csv("2019_NY_Weather.csv")
weather$TAVG <- (weather$TMAX + weather$TMIN)/2
```

#### Clean and prepare dataset for analysis
```{r}
# Transform variables
citisample$start.station.id <- as.factor(citisample$start.station.id)
citisample$end.station.id <- as.factor(citisample$end.station.id)
citisample$bikeid <- as.factor(citisample$bikeid)
citisample$usertype <- as.factor(citisample$usertype)
citisample$gender <- as.factor(citisample$gender)

# Extract some data from the date column 
# Since the format of each time is the same, we can get day and month by extracting based on character position 
citisample$day <- as.numeric(str_sub(as.character(citisample$starttime),9,10))
citisample$month <- as.numeric(str_sub(as.character(citisample$starttime),6,7))
# Put the day and month together to get dates, and we are only using 2019 data so hard code 2019
citisample$date <- paste(citisample$month, "/", citisample$day, "/19",  sep = "")
citisample$date <- as.Date(citisample$date, format = "%m/%d/%y")
# Extract weekend/weekday
citisample$DoW <- format(citisample$date, "%u")
citisample$dayid <- ifelse(citisample$DoW < 6, "Weekday", "Weekend")
citisample$dayid <- as.factor(citisample$dayid)
# Create another column for merging bike and weather data
citisample$mergedate <- paste(citisample$month, "/", citisample$day, sep = "")
# Lastly, format day and month as factors for visualizations
citisample$day <- as.factor(citisample$day)
citisample$month <- as.factor(citisample$month)
# Also, extract hour from the starttime
citisample$hour <- as.numeric(str_sub(as.character(citisample$starttime), 12, 13))
citisample$hour <- as.factor(citisample$hour)
# Create a column for approximate age, and one for age group.
citisample$age <- 2020 - citisample$birth.year
citisample$agegroup <- ifelse(citisample$age < 20, "<20", ifelse(citisample$age < 40, "20-40", "40+"))
```

```{r}
# Calculating the distance (mi) travelled for each ride:
R = 3958  # radius of earth in miles (at New York's latitude)
citisample$distance <- distHaversine(cbind(citisample$start.station.longitude, citisample$start.station.latitude), cbind(citisample$end.station.longitude, citisample$end.station.latitude), R)

# Fixing the date in the weather data so it matches the merge column in the bike data
weather$DATE <- str_sub(weather$DATE, 1, nchar(as.character(weather$DATE))-5)

```

#### Create a table with data grouped by date
```{r}
avgdata <- citisample %>%
  group_by(mergedate) %>%
  summarize(count = n(),
            dist = mean(distance, na.rm = TRUE),
            dur = mean(tripduration, na.rm = TRUE),
            speed = dist/dur)
```

#### Merge weather data and avg by date data
```{r}
combined_avg <- merge(avgdata, weather, by.x = "mergedate", by.y = "DATE")
```

#### Making graphs to show the relationships between different weather conditions and citibike data
#### First, we look at impact of temperature on number of rides
```{r}
# Impact of temp (Took an average)
baseplot <- ggplot(data = combined_avg, aes(x = TAVG, y = count))
baseplot + geom_line() + geom_smooth()
```
Here we see that the number of bike rides follows a general upward trend until an average temperature of about 74 degrees, then the number of rides starts to decrease. We can also separate these averages by variables such as weekday/weekend and gender to see if there is a difference in the effect. 

##### Looking at how number of rides is affected by temp during the weekdays/weekends. 
```{r}
# Seperate the avg data by weekday/weekend
avgdata <- citisample %>%
  group_by(mergedate, dayid) %>%
  summarize(count = n(),
            dist = mean(distance, na.rm = TRUE),
            dur = mean(tripduration, na.rm = TRUE),
            speed = dist/dur)
combined_Week <- merge(avgdata, weather, by.x = "mergedate", by.y = "DATE")

# Make a set of plots
baseplot_Week <- ggplot(data = combined_Week, aes(x = TAVG, y = count))
baseplot_Week + geom_line() + geom_smooth() + facet_wrap(~ dayid)
```
We see that weekday and weekend look pretty similar, except that the weekend graph has a steeper slope going up to the peak at 74 degrees. This makes sense because during the weekend, people who are using bikes are more likely using them for leisure, and so they can easily decide not to use a bike if the temperature is not perfect. However, during the weekday, many bike riders are riding to work, and they have to use a bike no matter the temperature. 

##### Looking at how number of rides is affected by temp based on gender. 
```{r}
avgdata <- citisample %>%
  group_by(mergedate, gender) %>%
  summarize(count = n(),
            dist = mean(distance, na.rm = TRUE),
            dur = mean(tripduration, na.rm = TRUE),
            speed = dist/dur)
combined_gender <- merge(avgdata, weather, by.x = "mergedate", by.y = "DATE")

# Make a set of plots
baseplot_gender <- ggplot(data = combined_gender, aes(x = TAVG, y = count))
baseplot_gender + geom_line() + geom_smooth() + facet_wrap(~ gender)
```
Here we run into the problem of having a large difference in ride numbers across genders, so it is hard to compare the graphs, but in general we can still see that the same curve of having the most number of rides at around 75 degrees is present across all gender categories. 

##### Looking at how number of rides is affected by temp based on usertype. 
```{r}
avgdata <- citisample %>%
  group_by(mergedate, usertype) %>%
  summarize(count = n(),
            dist = mean(distance, na.rm = TRUE),
            dur = mean(tripduration, na.rm = TRUE),
            speed = dist/dur)
combined_userT <- merge(avgdata, weather, by.x = "mergedate", by.y = "DATE")

# Make a set of plots
baseplot_userT <- ggplot(data = combined_userT, aes(x = TAVG, y = count))
baseplot_userT + geom_line() + geom_smooth() + facet_wrap(~ usertype)
```
The same problem as before is here, where the number of total bikes is very different between customers and subscribers, but we can again see that the general curve is the same. 

#### Next, we will look at how temperature affects average speed of rides. 
```{r}
# Impact of temperature
baseplot <- ggplot(data = combined_avg, aes(x = TAVG, y = speed))
baseplot + geom_line() + geom_smooth()
```
We can see that average speed differs a lot, but the general trend is that average speed decreases as temperatures increase. 

##### We can also see if different variables alter this relationship the same way we did with the affect of temperature on number of rides. 
```{r}
# Make a set of plots based on gender 
baseplot_gender <- ggplot(data = combined_gender, aes(x = TAVG, y = speed))
baseplot_gender + geom_line() + geom_smooth() + facet_wrap(~ gender)
```
This graph shows that on average, gender 0 is slower than the other genders, but the overall effect of increasing temperatures on speed does not change; there is still a slight decrease in speed as temperature rises. There is also a notable extra dip in the average speed of gender 0 around 74 degrees, which is interesting because that was also the peak in terms of number of rides. 

```{r}
# Make a set of plots based on dayid
baseplot_Week <- ggplot(data = combined_Week, aes(x = TAVG, y = speed))
baseplot_Week + geom_line() + geom_smooth() + facet_wrap(~ dayid)
```
During the weekdays, average speed declines slightly with temperature. During the weekend, average speed is at a maximum when it is around 40 degrees, and a minimum when it is at 65 degrees. 


#### Next we look at the impact of wind speed on bike data
```{r}
# Impact of Wind Speed
baseplot <- ggplot(data = combined_avg, aes(x = AWND, y = count))
baseplot + geom_line() + geom_smooth()
```
The peak number of rides occurs when wind speed is 2.5, which suggests that people enjoy biking when there is a bit of a breeze outside. However, anything more than that and the number of rides begins to decrease steadily as wind increases. 

#### We can also look at the impact of rainfall on bike data

```{r}
# Impact of Rain
baseplot <- ggplot(data = combined_avg, aes(x = PRCP, y = count))
baseplot + geom_line() + geom_smooth()
```

```{r}
# Impact of Snowfall
baseplot <- ggplot(data = combined_avg, aes(x = SNOW, y = count))
baseplot + geom_line() 
```

```{r}
# Impact of Snow Depth
baseplot <- ggplot(data = combined_avg, aes(x = SNWD, y = count))
baseplot + geom_line()
```


