---
title: "Group Project TO404"
author: "Lena Sophie Ivens"
date: "11/28/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Import and Consolidate Data
```{r, cache=TRUE}
# load library dplyr (for data manipulation)
library(dplyr)
# read CSV file containing 5% sample
sample19 <- read.csv("~/Downloads/sample19.csv")
# convert fields to factors as needed
sample19$starttime <- as.factor(sample19$starttime)
sample19$start.station.name <- as.factor(sample19$start.station.name)
sample19$end.station.name <- as.factor(sample19$end.station.name)
sample19$usertype <- as.factor(sample19$usertype)
sample19$birth.year <- as.factor(sample19$birth.year)
sample19$gender <- as.factor(sample19$gender)
```

Most traveled bikes - need maintenance

### Most travelled bikes
```{r}
# load library "hans" (for calculating distance)
library(hans)
# make column that specifies distance per trip in mi
sample19$distance <- haversine(sample19$start.station.latitude, sample19$start.station.longitude, sample19$end.station.latitude, sample19$end.station.longitude)*0.621371
# load library "ggplot" (for data visualization)
library(ggplot2)
# group sample by bikeid and distance
bikes <- sample19 %>%
        group_by(bikeid) %>%
        mutate(distance_per_bike = sum(distance)) %>%
        mutate(trips_per_bike = n())
# remove duplicate bikeid's and create new data frame
bikes1 <- select(bikes, bikeid, distance_per_bike, trips_per_bike)
bikes1 <- bikes1[!duplicated(bikes1$bikeid), ]
# create a scatter plot with distance and number of trips
plot1 <- ggplot(data = bikes1, aes(x = trips_per_bike, y = distance_per_bike)) + geom_point()
plot1
# identifying most used and least used bikes
mostusedbikes <- head(bikes1[order(bikes1$distance_per_bike, decreasing = TRUE),], n = 100)
mostusedbikes
leastusedbikes <- tail(bikes1[order(bikes1$distance_per_bike, decreasing = TRUE),], n = 100)
leastusedbikes
```

- Some bikes have been used more than others ultimately requiring more maintenance. As we can see the distance traveled ranges from `r round(max(bikes1$distance_per_bike),2)` miles to 0 miles. The scatterplot reveals that there is no unusual pattern since the more a bike is booked the larger the distance traveled. 
